---
title: "COVID Fund Spending Analysis"
subtitle: "Application Exercise"
---

## Learning objective

By the end of this exercise, you will reproduce and extend the **COVID recovery fund scatter plot** from the slides, analyzing the relationship between student poverty and ESSER spending in a state of your choice.

:::{.callout-note title="Sharing your work"}
Save your favorite plot using the naming convention `firstname_analysis_description.png` and upload it to the [shared Google Drive folder](https://drive.google.com/drive/folders/1JwmVtz0OA7nOz89DOHGzH3gddiG5VR_a?usp=drive_link).
:::

## Part 1: Setup

Load your libraries and pick a state. We'll start with a **single year** to explore the data before scaling up to multi-year analysis.

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(edfinr)
library(scales)
library(viridis)

# TODO: Change these to your name and state
your_name <- "alex"
my_state <- "GA"
```

Pull a single year of full data for your state:

```{r}
#| label: pull-data
state_data <- get_finance_data(
  yr = "2022",
  geo = my_state,
  dataset_type = "full"
)
```

Take a quick look at what you have:

```{r}
#| label: explore
glimpse(state_data)
```

## Part 2: Select and clean

The full dataset has hundreds of columns. Replicate the `select()` call from the slides, picking the variables we need for COVID spending analysis:

```{r}
#| label: select-vars
#| eval: false
covid_exp <- state_data |>
  select(
    # identifiers and enrollment
    ncesid, year, state, cbsa, cong_dist, dist_name, enroll,
    # revenue per-pupil
    ___,
    # socio-economic variables
    ___,
    # expenditure data
    ___
  )
```

:::{.callout-tip collapse="true" title="Solution"}
```{r}
#| label: select-vars-solution
#| eval: false
covid_exp <- state_data |>
  select(
    # identifiers and enrollment
    ncesid, year, state, cbsa, cong_dist, dist_name, enroll,
    # revenue per-pupil
    rev_total_pp, rev_local_pp, rev_state_pp, rev_fed_pp,
    # socio-economic variables
    mhi, mpv, ba_plus_pct, stpov_pct, urbanicity,
    # expenditure data
    exp_cur_total, exp_covid_total, exp_covid_instr
  )
```
:::

## Part 3: Per-pupil calculations

The expenditure columns are **totals**, not per-pupil. Use `across()` to convert all of them at once, then calculate the instructional share of COVID spending:

```{r}
#| label: per-pupil
#| eval: false
covid_exp <- covid_exp |>
  mutate(
    # convert all exp_ columns to per-pupil
    across(
      ___,
      ___,
      .names = "___"
    ),
    # what share of COVID funds went to instruction?
    exp_covid_instr_pct = ___
  )
```

This is the key `across()` pattern from the slides:

- `starts_with("exp_")` selects all expenditure columns
- The lambda `\(x) x / enroll` divides each by enrollment
- `.names = "{.col}_pp"` creates new columns with a `_pp` suffix

:::{.callout-tip collapse="true" title="Solution"}
```{r}
#| label: per-pupil-solution
#| eval: false
covid_exp <- covid_exp |>
  mutate(
    # convert all exp_ columns to per-pupil
    across(
      starts_with("exp_"),
      \(x) x / enroll,
      .names = "{.col}_pp"
    ),
    # what share of COVID funds went to instruction?
    exp_covid_instr_pct = exp_covid_instr / exp_covid_total
  )
```
:::

## Part 4: Build the scatter plot

ESSER funds were allocated via **Title I formulas**. Did higher-poverty districts in your state actually spend more? Build a scatter plot to find out.

Fill in the aesthetics and layers:

```{r}
#| label: scatter-plot
#| eval: false
ggplot(data = covid_exp) +
  geom_point(
    aes(
      x = ___,
      y = ___,
      color = ___,
      size = ___
    ),
    alpha = .6
  ) +
  geom_smooth(
    aes(
      x = ___,
      y = ___
    ),
    method = "lm", color = "orange"
  ) +
  scale_x_continuous(labels = label_percent()) +
  scale_y_continuous(labels = label_dollar()) +
  scale_color_viridis(labels = label_percent()) +
  scale_size_area(max_size = 10, labels = label_comma()) +
  theme_bw() +
  labs(
    x = "Student Poverty %",
    y = "Total COVID Fund Expenditures Per-Pupil",
    size = "Enrollment",
    color = "Instructional Exp. as % of\nTotal COVID Fund Exp.",
    caption = "Data Source: edfinr"
  )
```

What to look for:

- **Slope direction**: Do higher-poverty districts spend more COVID funds per pupil?
- **Color gradient**: How much variation is there in the instructional share?
- **Size**: Do larger districts behave differently than smaller ones?

:::{.callout-tip collapse="true" title="Solution"}
```{r}
#| label: scatter-plot-solution
#| eval: false
ggplot(data = covid_exp) +
  geom_point(
    aes(
      x = stpov_pct,
      y = exp_covid_total_pp,
      color = exp_covid_instr_pct,
      size = enroll
    ),
    alpha = .6
  ) +
  geom_smooth(
    aes(
      x = stpov_pct,
      y = exp_covid_total_pp
    ),
    method = "lm", color = "orange"
  ) +
  scale_x_continuous(labels = label_percent()) +
  scale_y_continuous(labels = label_dollar()) +
  scale_color_viridis(labels = label_percent()) +
  scale_size_area(max_size = 10, labels = label_comma()) +
  theme_bw() +
  labs(
    x = "Student Poverty %",
    y = "Total COVID Fund Expenditures Per-Pupil",
    size = "Enrollment",
    color = "Instructional Exp. as % of\nTotal COVID Fund Exp.",
    caption = "Data Source: edfinr"
  )
```
:::

## Part 5: Extend to multi-year (stretch goal)

In the slides, we used **range syntax** to pull multiple years and `facet_wrap()` to compare them. Try it for your state.

Pull multi-year data and apply the same pipeline:

```{r}
#| label: multi-year-pull
#| eval: false
state_multi <- get_finance_data(
  yr = "2020:2022",
  geo = my_state,
  dataset_type = "full"
)
```

```{r}
#| label: multi-year-clean
#| eval: false
# Apply the same select → across → mutate pipeline
covid_multi <- state_multi |>
  select(
    # TODO: same columns as Part 2
  ) |>
  mutate(
    across(
      # TODO: same per-pupil calculation as Part 3
    ),
    exp_covid_instr_pct = exp_covid_instr / exp_covid_total
  )
```

Now add `facet_wrap(~year)` to your scatter plot to see how COVID spending evolved from 2020 to 2022:

```{r}
#| label: multi-year-plot
#| eval: false
# TODO: Copy your scatter plot from Part 4 and add:
# - Change data to covid_multi
# - Add facet_wrap(~year) before theme_bw()
```

:::{.callout-tip collapse="true" title="Solution"}
```{r}
#| label: multi-year-solution
#| eval: false
covid_multi <- state_multi |>
  select(
    ncesid, year, state, cbsa, cong_dist, dist_name, enroll,
    rev_total_pp, rev_local_pp, rev_state_pp, rev_fed_pp,
    mhi, mpv, ba_plus_pct, stpov_pct, urbanicity,
    exp_cur_total, exp_covid_total, exp_covid_instr
  ) |>
  mutate(
    across(
      starts_with("exp_"),
      \(x) x / enroll,
      .names = "{.col}_pp"
    ),
    exp_covid_instr_pct = exp_covid_instr / exp_covid_total
  )

ggplot(data = covid_multi) +
  geom_point(
    aes(
      x = stpov_pct,
      y = exp_covid_total_pp,
      color = exp_covid_instr_pct,
      size = enroll
    ),
    alpha = .6
  ) +
  geom_smooth(
    aes(
      x = stpov_pct,
      y = exp_covid_total_pp
    ),
    method = "lm", color = "orange"
  ) +
  scale_x_continuous(labels = label_percent()) +
  scale_y_continuous(labels = label_dollar()) +
  scale_color_viridis(labels = label_percent()) +
  scale_size_area(max_size = 10, labels = label_comma()) +
  facet_wrap(~year) +
  theme_bw() +
  labs(
    x = "Student Poverty %",
    y = "Total COVID Fund Expenditures Per-Pupil",
    size = "Enrollment",
    color = "Instructional Exp. as % of\nTotal COVID Fund Exp.",
    caption = "Data Source: edfinr"
  )
```
:::

## Part 6: Extension challenge

Everyone should complete the scatter plot in Parts 4-5 first. Then pick **one** of the following extensions:

### Option A: Compare multiple states

Pull data for multiple states in one call using a comma-separated list in `geo`. How does the poverty-spending relationship differ across states?

```{r}
#| label: option-a
#| eval: false
# TODO: Update geo to pull multiple states using a comma list (no spaces)
# e.g., geo = "GA,TX,NC"
# Then facet by state (or facet_grid(state ~ year) for both dimensions)
```

:::{.callout-tip collapse="true" title="Option A Solution"}
```{r}
#| label: option-a-solution
#| eval: false
my_states <- "GA,TX,NC"

combined_raw <- get_finance_data(
  yr = "2020:2022",
  geo = my_states,
  dataset_type = "full"
)

combined <- combined_raw |>
  select(
    ncesid, year, state, cbsa, cong_dist, dist_name, enroll,
    rev_total_pp, rev_local_pp, rev_state_pp, rev_fed_pp,
    mhi, mpv, ba_plus_pct, stpov_pct, urbanicity,
    exp_cur_total, exp_covid_total, exp_covid_instr
  ) |>
  mutate(
    across(
      starts_with("exp_"),
      \(x) x / enroll,
      .names = "{.col}_pp"
    ),
    exp_covid_instr_pct = exp_covid_instr / exp_covid_total
  )

ggplot(data = combined) +
  geom_point(
    aes(
      x = stpov_pct,
      y = exp_covid_total_pp,
      color = exp_covid_instr_pct,
      size = enroll
    ),
    alpha = .6
  ) +
  geom_smooth(
    aes(
      x = stpov_pct,
      y = exp_covid_total_pp
    ),
    method = "lm", color = "orange"
  ) +
  scale_x_continuous(labels = label_percent()) +
  scale_y_continuous(labels = label_dollar()) +
  scale_color_viridis(labels = label_percent()) +
  scale_size_area(max_size = 10, labels = label_comma()) +
  facet_grid(state ~ year) +
  theme_bw() +
  labs(
    x = "Student Poverty %",
    y = "Total COVID Fund Expenditures Per-Pupil",
    size = "Enrollment",
    color = "Instructional Exp. as % of\nTotal COVID Fund Exp.",
    caption = "Data Source: edfinr"
  )
```
:::

### Option B: Swap the variable

Replace `stpov_pct` on the x-axis with a different socio-economic variable. How does the story change?

```{r}
#| label: option-b
#| eval: false
# TODO: Replace stpov_pct with mhi (median household income) or
# ba_plus_pct (% adults with BA+). Update the x-axis label.
# Does the relationship flip?
```

:::{.callout-tip collapse="true" title="Option B Solution"}
```{r}
#| label: option-b-solution
#| eval: false
ggplot(data = covid_multi) +
  geom_point(
    aes(
      x = mhi,
      y = exp_covid_total_pp,
      color = exp_covid_instr_pct,
      size = enroll
    ),
    alpha = .6
  ) +
  geom_smooth(
    aes(
      x = mhi,
      y = exp_covid_total_pp
    ),
    method = "lm", color = "orange"
  ) +
  scale_x_continuous(labels = label_dollar()) +
  scale_y_continuous(labels = label_dollar()) +
  scale_color_viridis(labels = label_percent()) +
  scale_size_area(max_size = 10, labels = label_comma()) +
  facet_wrap(~year) +
  theme_bw() +
  labs(
    x = "Median Household Income",
    y = "Total COVID Fund Expenditures Per-Pupil",
    size = "Enrollment",
    color = "Instructional Exp. as % of\nTotal COVID Fund Exp.",
    caption = "Data Source: edfinr"
  )
```
:::

### Option C: Within-state geography

Use `cbsa` or `cong_dist` to facet by metro area or congressional district, showing within-state variation in COVID spending patterns.

```{r}
#| label: option-c
#| eval: false
# TODO: Filter to districts with enough enrollment to be meaningful (e.g., > 500)
# Use facet_wrap(~cbsa) or facet_wrap(~cong_dist) to show geographic variation
# You may need to filter to a subset of geographies that have enough districts
```

:::{.callout-tip collapse="true" title="Option C Solution"}
```{r}
#| label: option-c-solution
#| eval: false
# Filter to 2022 and districts with meaningful enrollment
geo_data <- covid_multi |>
  filter(year == 2022, enroll > 500, !is.na(cong_dist))

# Keep only congressional districts with 5+ districts for readable facets
cds_with_data <- geo_data |>
  count(cong_dist) |>
  filter(n >= 5) |>
  pull(cong_dist)

geo_data |>
  filter(cong_dist %in% cds_with_data) |>
  ggplot() +
  geom_point(
    aes(
      x = stpov_pct,
      y = exp_covid_total_pp,
      color = exp_covid_instr_pct,
      size = enroll
    ),
    alpha = .6
  ) +
  geom_smooth(
    aes(
      x = stpov_pct,
      y = exp_covid_total_pp
    ),
    method = "lm", color = "orange"
  ) +
  scale_x_continuous(labels = label_percent()) +
  scale_y_continuous(labels = label_dollar()) +
  scale_color_viridis(labels = label_percent()) +
  scale_size_area(max_size = 10, labels = label_comma()) +
  facet_wrap(~cong_dist) +
  theme_bw() +
  labs(
    x = "Student Poverty %",
    y = "Total COVID Fund Expenditures Per-Pupil",
    size = "Enrollment",
    color = "Instructional Exp. as % of\nTotal COVID Fund Exp.",
    caption = "Data Source: edfinr"
  )
```
:::

## Part 7: Save and share

Save your best plot and upload it to the shared folder:

```{r}
#| label: save-plot
#| eval: false

# paste this code right after your selected plot code
# run your plot code, then fill in your analysis description,
# then run the ggsave command - this will save the png of your
# plot to your current working directory.

# fill this in with a pithy snake_case description of your analysis
# example: analysis of kentucky and ohio covid expenditures might be
# "ky_oh_covid_exp"
analysis_description <- ""

ggsave(
  filename = paste0(your_name, "_", analysis_description, ".png"),
  units = "in",
  width = 10,
  height = 6
)
```

Upload your saved plot to the [shared Google Drive folder](https://drive.google.com/drive/folders/1JwmVtz0OA7nOz89DOHGzH3gddiG5VR_a?usp=drive_link).

### Quality checklist

Before you upload, check:

- Does the title tell the story?
- Are axes labeled clearly with `labs()`?
- Are dollar values formatted with `scales::label_dollar()`?
- Would a subtitle or caption add useful context?
